\begin{agradecimentos}

    First and foremost, I would like to thank my family—Raul, Barbara, Sandra, and Ester—for their unwavering support throughout this journey, especially my father, João Luiz Franco, who has always encouraged my studies. It is an honor to be graduating from the same Institute and in the same course where my hero graduated in 1987.

    I extend my deepest gratitude to my advisor, Filipe Alves Neto Verri, who has not only guided me this semester but has been a constant source of wisdom and inspiration throughout much of my academic journey. Thank you for your patience, guidance, encouragement and the opportunity you gave me, professor.

    My heartfelt thanks also go to Manoel Vilela Machado Neto, whose collaboration and friendship were instrumental in completing this work. Our numerous discussions provided invaluable learning experiences and opportunities for growth.

    I am equally grateful to the entire ICMC staff, especially the graduation service team—for their patience with my many requests—and to the library, as well as the cleaning and general services staff, whose kindness and support made ICMC feel like a second home. I owe thanks to all the professors who contributed directly to this project through their teaching, particularly in graph theory: professors Ricardo, Diego, Francisco, Rudinei, Cynthia, Luis, and João Batista.

    Furthermore, I also thank my IFSP professors Renato, André, Malu, Rodrigo, Fábio and Jorge, who encouraged me to get into college. I am sure that without their support, specially from professor Renato who passed countless days helping me in physics, I would not be here.

    To my examination board, I am deeply thankful for this opportunity. Professor Solange, the first member of the board, lectured me Artificial Intelligence, and I could learn so much from someone that always brought a positive outlook and saw the best in us. Maria Carolina Monard, who advised both Solange and my father, was instrumental in my father's journey and consequently mine. Though she is no longer with us, I dedicate this project to her in recognition of her incredible contributions to all our lives. It is also a privilege to have my dear friend André Fakhoury as the second member of the board. Thank you for everything, André—it's interesting that this project is about graphs, but you've been helping me with graphs even before I entered at USP, back when I would bother you in GEMA's group.

    Lastly, I thank the love of my life, Isis, for making me happier every day, for all the care and understanding during this period, my friends for their unwavering support, and God, who has never abandoned me.

\end{agradecimentos}
