% ----------------------------------------------------------
% Introdução
% ----------------------------------------------------------
\chapter[Introduction]{Introduction}
\label{Introdução}

\noindent The aviation industry increasingly relies on data-driven approaches to improve operational efficiency and reduce delays. Among the pressing challenges in air traffic management is the prediction of `holding' maneuvers, where aircraft are instructed to delay landing due to factors such as airport congestion, adverse weather, or airspace limitations. While holding patterns are necessary for safety, they contribute to increased fuel consumption, emissions, and passenger dissatisfaction. This project aims to enhance the accuracy of holding predictions using machine learning (ML) models based on graph-structured data, specifically employing advanced methodologies in Graph Machine Learning (Graph ML) and Graph Neural Networks (GNNs).

Traditional machine learning applications in aviation have primarily focused on flight delay prediction and air traffic flow management. For instance, delay predictions based on weather conditions, airport congestion, and flight schedules have been widely studied \cite{lambelho2020assessing, gui2019flight}. However, these models often rely on tabular data representations, which limit their ability to capture complex relational patterns among airports and other influencing factors. Additionally, research specifically focused on holding maneuvers is limited and generally lacks machine learning and network-based approaches that can model the spatial and temporal dependencies intrinsic to air traffic data \cite{lee2020development, smith2008management}.

The use of graph-based machine learning methods is rapidly advancing in the field of intelligent transportation, where graph structures effectively capture complex spatial and temporal relationships across networks. A recent survey on GNNs in intelligent transportation systems,  \citeonline{gnn_survey_transportation}, highlights their application across a variety of domains, including traffic forecasting, demand prediction, and urban planning. This survey underscores the power of GNNs in applications where data is inherently interconnected, such as autonomous vehicle routing and intersection management. By organizing studies within these domains, they identify distinct opportunities and challenges, particularly in multi-modal models and reinforcement learning applications. Similarly, \citeonline{zhao2020tgcn} demonstrate the value of GNNs in the specific context of real-time traffic forecasting with their T-GCN (Temporal Graph Convolutional Network) model. By combining Graph Convolutional Networks (GCN) and Gated Recurrent Units (GRU), the T-GCN model captures both spatial and temporal dependencies, achieving state-of-the-art accuracy in urban traffic prediction tasks. These studies exemplify the increasing role of GNNs in transportation-related decision-making, showing potential for improved accuracy and efficiency in complex, dynamic systems.

The project employs two main approaches:

\begin{itemize}
    \item \textbf{Tabular-based approach:} We use the CatBoost model, leveraging graph features—such as centrality and connectivity metrics—that capture the significance of directed edges (flights) within the network \cite{prokhorenkova2018catboost}.
    \item \textbf{Graph Attention Network (GAT) approach:} We compare with GATs that have proven effective in applications where relational data is essential, making them a promising choice for capturing the interconnected nature of air traffic \cite{velickovic2017graph}.
\end{itemize}

The contributions of this project are twofold. First, we introduce a novel application of graph-based ML models for predicting holding events, providing a refined view of airport interdependencies. Second, by comparing CatBoost with the GAT model, we evaluate which approach better captures the graph topology of the air traffic and have a better predictive performance. This research has potential implications for improved fuel efficiency, reduced delays, and enhanced passenger experiences by refining model selection and feature engineering strategies tailored to aviation applications. For the best of our knowledge there is no previous work that approached the problem of flight delay due to holding maneuver using Graph ML.

The structure of this work is as follows. In Section \ref{TheoreticalFramework}, we review relevant literature and build the theoretical background needed. Section \ref{Materials} details the dataset and the modelling. Section \ref{Results} discusses the experimental setup, with results and comparison analysis presented and model deployment. Finally, Section \ref{Conclusion} provides concluding remarks about my Computer Science Bachelor's.
