% ---
% Conclusão
% ---
\chapter{Conclusion}
\label{Conclusion}

\section{Project Contributions to the Student}

This project was an excellent way to reinforce the knowledge I gained in my undergraduate program, particularly in areas like algorithms and graph applications. Working at ICMC was an invaluable opportunity to engage with experts in graph machine learning and complex networks. Through this project, I gained insight into both fields—a feat that would have been challenging without the foundation provided by courses like Complex Networks and Artificial Intelligence, which taught me the basics of each. Additionally, this experience provided an ideal introduction to studying \emph{graph neural networks}, a field I plan to pursue further.

\section{Relationship between the Undergraduate Course and the Project}

The undergraduate curriculum strongly aligned with this project. Key subjects included Algorithms and Data Structures, Computational Modeling in Graphs, Artificial Intelligence, High-Performance Computing, Linear Algebra, Calculus, Stochastic Processes, Statistics, and Advanced Algorithms and Applications. Together, these courses provided a comprehensive foundation in computer science, with ample focus on graph theory, machine learning, and algorithm design. The curriculum thus offered a robust theoretical framework that enriched my understanding of complex networks and graph-based models.

Another crucial aspect of my program was the opportunity to participate in teaching and research. This experience not only deepened my technical knowledge but also honed my communication skills, enabling me to articulate ideas and solutions more effectively.

\section{Reflections on the Undergraduate Course}

The Bachelor of Computer Science at the Instituto de Ciências Matemáticas e de Computação, Universidade de São Paulo, has been instrumental in providing me with knowledge, friendships, determination, and joy. I am deeply grateful for the supportive structure at ICMC, which offered everything necessary for personal and academic growth. Being surrounded by dedicated and inquisitive peers and professors has been the highlight of my time here, creating a stimulating environment that constantly pushed me to expand my horizons as a person, a scholar, and a future professional.

That said, I do have one critique regarding the course curriculum. Recently, it feels as though the Bachelor of Computer Science program has shifted its focus more toward the job market than foundational science. While I understand the need to prepare students for industry, I believe that we must preserve the course's scientific roots. For instance, `Research Methodology' is not a required course, yet it is fundamental for academic development, while other subjects more geared toward job readiness are mandatory. The removal of `Calculus 4` from the curriculum further illustrates this shift away from theoretical depth, despite its relevance in areas like machine learning.

To address this, I would suggest offering two distinct tracks within the program: an Academic Track and a Professional Track. This structure would allow students interested in research and theoretical science to focus on these areas, while those aiming for immediate industry applications could focus on practical skills. This change might also encourage more students to pursue advanced degrees at ICMC by fostering a stronger scientific foundation during their undergraduate studies.

Despite these critiques, I am confident that choosing this program was one of the best decisions I have made. The course has opened doors to opportunities I had not even dreamed of achieving.
